raptr: Representative and Adequate Prioritisation Toolkit in R

### 2016-07-25 code/rmarkdown/article.Rmd
+ assorted manuscript fixes

"
output:
  rmarkdown::pdf_document:
    toc: false
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: preamble-latex.tex
fontsize: 11pt
documentclass: article
bibliography: references.bib
csl: reference-style.csl
---

@@

<!--  title page -->
Short running title: Representative and adequate prioritisations \newline
Abstract word count: 350 / 350  
Total word count: XXXXX / 8000  
Number of references: 37
Number of figures: 8
Number of tables: 0  
Number of textboxes: 0  
Number of equations: 8  
\newpage

@@

To illustrate these concepts, consider an example conservation planning scenario example involving an attribute space and demand points. We wish to develop a prioritisation for a single species that has four populations in the study area. However, we can only afford to preserve three populations. We aim preserve the adaptive landscape of the species, and to achieve this, we will preserve populations inhabiting different environmental conditions. To describe environmental variation, we obtain data on the environmental conditions (rainfall (mm) and temperature ($^{\circ}$C)) where each population is found. These environmental data are then used to construct a two-dimensional environmental attribute space. Next, we generated demand points as equi-distant points between the range of values where the populations were found. By comparing the distribution of the demand points to the distribution of the populations in the attribute space, we can identify a suitable prioritisation (Fig. 1). We can see that preserving both populations $A$ and $C$, effectively "doubles-up" on the same environmental characteristics, resulting in considerable redundancy in the solution. Instead, a more representative sample of the intra-specific variation could be captured by securing populations $A$, $B$, and $D$. This example demonstrates how the inclusion of biodiversity processes can guide the reserve selection process.

The formulations used to express the reserve selection problem in the \texttt{raptr R} package are based on a combination of the \texttt{Marxan} reserve selection problem and the uncapacitated facility location problems [@r451]. Although the \texttt{raptr R} package provides two novel formulations, for brevity, we will define the simpler formulation--referred to as the unreliable formulation--below and define the more complex version--the reliable formulation--in the Supporting Information S1. These formulations are named after the unreliable and reliable facility location formulations from which they are based upon [@r16]. The key difference between these two formulations is that the reliable formulation explicitly considers the probability that planning units are occupied when calculating the proportion of an attribute space sampled in a solution. All mathematical terms defined hereafter are described in Table S1. For convenience, the cardinality of sets will be denoted using the same symbol used to denote the variable.

@@

## OPTIMISATION
The unreliable formulation is non-linear. However, the non-linear components can be linearised using existing techniques. The expression $X_j X_k$ in (eqn 3a) can be linearised using methods described by Beyer et al. [-@r426]. Linearised versions of the problems can be solved using commercial exact algorithm solvers. The \texttt{raptr} R package provides functions to express conservation planning data as optimisation problems using linearised versions of the unreliable and reliable formulations. These optimisation problems can then be solved to generate prioritisations using the commercial \texttt{Gurobi} software suite (\url{http://www.gurobi.com}). Presently, academics can obtain a license at no cost from the Gurobi website. After installing the \texttt{Gurobi} software suite, users will need to install the \texttt{Gurobi R} package.

@@

The uniform species was simulated to occur with a constant probability of occupancy across the study area (Fig. 2a). The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount-based targets (Fig. 3a) selected `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units near the southern end of study area. This configuration is an artifact of the method used to solve this particular instance of the reserve selection problem. In fact, for this species, all prioritisations containing `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units are optimal when considering only `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. In the absence of criteria to guarantee representativeness, reserve selection methods may or may not return solutions that secure a representative sample of the features. In this particular case, the solution does not secure a representative sample of the species' range (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled).

@@

## CASE STUDY 2
### Methods
Here we used space-based targets to generate a prioritisation securing a representative sample of a species' intra-specific genetic variation. We used species occurrence and multilocus AFLP data collected by the international IntraBioDiv project in the European Alps [@r462; see @r461 and @r463 for further explanation of data collection methods]. Althougthis dataset contains multiple plant species, we used data for the `r tolower(cs2.params.LST[[MODE]]$common.name)` (\textit{`r gsub('_', ' ',cs2.params.LST[[MODE]]$species.name,fixed=TRUE)`}), a wide-spread species with significant genetic structure [@r454]. Members of the IntraBioDiv project collected data using a $20^{\prime}$ longitude by $21^{\prime}$ latitude grid (approx. 22.3 km $\times$ 25 km; Fig. 7a). They visited every second grid cell, and if the `r tolower(cs2.params.LST[[MODE]]$common.name)` was detected in a cell, samples were collected from three individuals. Samples were genotyped using amplified fragment length polymorphisms [AFLP; @r453], and used to construct matrices denoting the presence/absence of polymorphisms at loci. In total, `r structurer::n.samples(cs2.spp.StructureData.LST[[1]])` individuals were genotyped at `r structurer::n.loci(cs2.spp.StructureData.LST[[1]])` markers.

@@

The solution generated using just the amount target failed to preserve a representative portion of the species' genetic variation (`r round(cs2.spp.DF$genetic[1],2)` % sampled; Figs. 8a and 8c). This solution only sampled planning units that contained individuals belonging to one of the main two groups. On the other hand, the solution generated using additional genetic targets selected planning units belonging to both of the main groups. Note that since these solutions were generated using opportunity cost data, the solution generated using only the amount target is essentially the minimum number of least costly areas needed to fulfill the amount target. The only difference between the two solutions is a single planning unit. By swapping this single planning unit, the solution generated using amount and genetic targets was able to preserve a representative portion of the species' genetic variation `r round(cs2.spp.DF$genetic[1],2)` for only a minor increase in cost (`r round(cs2.spp.DF$Score[1],2)` total cost compared to `r round(cs2.spp.DF$Score[2],2)` total cost).

# Implications and future directions
The \texttt{raptr} R package provides a unified approach to reserve selection. One of the key advantages of this package is that it is general enough to incorporate any spatial variable as an attribute space. This package can accommodate intrinsic or extrinsic variation to the feature(s). For example, adaptation processes could be secured using environmental variation or genetic variation among loci under selection. Ecological processes, such as predator-prey interactions, could be secured by capturing intra-specific trophic variation. Different attribute spaces with different targets can be also specified for different species. As long as the variation can be described using Euclidean distances, the \texttt{R} package can be used to obtain a representative sample (note that data may require transformation to conform to this assumption). This \\texttt{R} package provides decision makers with the tools to generate prioritisations that secure both biodiversity patterns and the processes that maintain biodiversity. Additionally, the package contains functionality to accommodate uncertainty in the distribution of features, and also identify suitably connected reserves. Both the simulated and case study species suggest that conservation planning exercises need to explicitly consider biodiversity processes during the reserve selection process to ensure they are captured.

@@

The formulation requires spatially comprehensive data to map attribute spaces to planning units. For instance, in both case-studies, all planning units occupied by the species were associated with values/coordinates in the attribute spaces. However, most real-world data sets are patchy--some planning units will be occupied by species for which attribute space data is not available. To use such patchy data with the \\texttt{raptr R} package, the gaps in the data must first be filled in as a pre-processing step. Spatially explicit models could be used to estimate values in locations that are missing from a data set [eg. krigging or generalised dissimilarity models; @r467; @r466]. This approach has been successfully applied to a range of biological data sets [@r468].

To maximise the long-term persistence of biodiversity, decision makers need to identify prioritisations that preserve existing patterns of biodiversity and the processes that support them. To achieve this, conservation planners need a decision support tool that can explicitly accommodate biodiversity patterns and processes. Here, we developed the \texttt{raptr R} package to fill this void. By exploring the functionality of this package using several simulated species, we found that including space-based targets can radically change a prioritisation for the simplest of species.

@@

# Data accessibility
All data, code, and results are stored in an online repository (\url{www.github.com/paleo13/raptr-manuscript}) to permit replication and validation of this study.

>>>>

"
output:
  rmarkdown::pdf_document:
    toc: false
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: preamble.tex
fontsize: 11pt
documentclass: article
bibliography: references.bib
csl: reference-style.csl
---

@@

<!--  title page -->
Short running title: Representative and adequate prioritisations \newline
Abstract word count: 348 / 350  
Total word count: 5481 / 7000
Number of references: 37
Number of figures: 8  
Number of tables: 0  
Number of textboxes: 0  
Number of equations: 8  
\newpage

@@

To illustrate these concepts, consider an example conservation planning scenario example involving an attribute space and demand points. We wish to develop a prioritisation for a single species that has four populations in the study area. However, we can only afford to preserve three populations. We aim preserve the adaptive landscape of the species, and to achieve this, we will preserve populations inhabiting different environmental conditions. To describe environmental variation, we obtained data on the environmental conditions (rainfall (mm) and temperature ($^{\circ}$C)) where each population is found. These environmental data are then used to construct a two-dimensional environmental attribute space. Next, we generated demand points as equi-distant points between the range of values where the populations were found. By comparing the distribution of the demand points to the distribution of the populations in the attribute space, we can identify a suitable prioritisation (Fig. 1). We can see that preserving both populations $A$ and $C$, effectively "doubles-up" on the same environmental characteristics, resulting in considerable redundancy in the solution. Instead, a more representative sample of the intra-specific variation could be captured by securing populations $A$, $B$, and $D$. This example demonstrates how the inclusion of biodiversity processes can guide the reserve selection process.

The formulations used to express the reserve selection problem in the \texttt{raptr R} package are based on a combination of the \texttt{Marxan} reserve selection problem and the uncapacitated facility location problems [@r451]. Although the \texttt{raptr R} package utlises two novel formulations, for brevity, we will define the simpler formulation--referred to as the unreliable formulation--below and define the more complex version--the reliable formulation--in the Supporting Information S1. These formulations are named after the unreliable and reliable facility location formulations from which they are based upon [@r16]. The key difference between these two formulations is that the reliable formulation explicitly considers the probability that planning units are occupied when calculating the proportion of an attribute space sampled in a solution. All mathematical terms defined hereafter are described in Table S1. For convenience, the cardinality of sets will be denoted using the same symbol used to denote the variable.

@@

## OPTIMISATION
The unreliable formulation is non-linear. However, the non-linear components can be linearised using existing techniques. The expression $X_j X_k$ in (eqn 3a) can be linearised using methods described by Beyer et al. [-@r426]. Linearised versions of the problems can be solved using commercial exact algorithm solvers. The \texttt{raptr} R package provides functions to express conservation planning data as optimisation problems using linearised versions of the unreliable and reliable formulations. These optimisation problems can then be solved to generate prioritisations using the commercial \texttt{Gurobi} software suite (\url{http://www.gurobi.com}). Presently, academics can obtain a license at no cost from the Gurobi website.

@@

The uniform species was simulated to occur with a constant probability of occupancy across the study area (Fig. 2a). The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount-based targets (Fig. 3a) selected `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units near the southern end of study area. This configuration is an artifact of the method used to solve this particular instance of the reserve selection problem. In fact, for this species, all prioritisations containing `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units are optimal when considering only `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. In the absence of criteria to guarantee representativeness, reserve selection methods may or may not return solutions that secure a representative sample of the features. In this particular case, the solution does not secure a representative sample of the species' range (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled). Note that the proportion here is negative because it performs worse than simply citing a planning unit in the centroid of the species' demand points.

@@

## CASE STUDY 2
### Methods
Here we used space-based targets to generate a prioritisation securing a representative sample of a species' intra-specific genetic variation. We used species occurrence and multilocus AFLP data collected by the international IntraBioDiv project in the European Alps [@r462; see @r461 and @r463 for further explanation of data collection]. Although this dataset contains multiple plant species, we used data for the `r tolower(cs2.params.LST[[MODE]]$common.name)` (\textit{`r gsub('_', ' ',cs2.params.LST[[MODE]]$species.name,fixed=TRUE)`}), a wide-spread species with significant genetic structure [@r454] as a simple case study. Members of the IntraBioDiv project collected data using a $20^{\prime}$ longitude by $21^{\prime}$ latitude grid (approx. 22.3 km $\times$ 25 km; Fig. 7a). They visited every second grid cell, and if the `r tolower(cs2.params.LST[[MODE]]$common.name)` was detected in a cell, samples were collected from three individuals. Samples were genotyped using amplified fragment length polymorphisms [AFLP; @r453], and used to construct matrices denoting the presence/absence of polymorphisms at loci. In total, `r structurer::n.samples(cs2.spp.StructureData.LST[[1]])` individuals were genotyped at `r structurer::n.loci(cs2.spp.StructureData.LST[[1]])` markers.

@@

The solution generated using just the amount target failed to preserve a representative portion of the species' genetic variation (`r round(cs2.spp.DF$genetic[1],2)` % sampled; Figs. 8a and 8c). This solution only sampled planning units that contained individuals belonging to one of the main two groups. On the other hand, the solution generated using additional genetic targets selected planning units belonging to both of the main groups. Note that since these solutions were generated using opportunity cost data, the solution generated using only the amount target is essentially the minimum number of least costly areas needed to fulfill the amount target. The only difference between the two solutions is a single planning unit. By swapping this single planning unit, the solution generated using amount and genetic targets was able to preserve a representative portion of the species' genetic variation (`r round(cs2.spp.DF$genetic[2],2)` % sampled; Figs. 8b and 8d) for only a minor increase in cost (`r round(cs2.spp.DF$Score[1],2)` total cost compared to `r round(cs2.spp.DF$Score[2],2)` total cost).

# Implications and future directions
The \texttt{raptr} R package provides a unified approach to reserve selection. One of the key advantages of this package is that it is general enough to incorporate any spatial variable as an attribute space. This package can accommodate intrinsic or extrinsic variation to the feature(s). For example, adaptation processes could be secured using environmental variation or genetic variation among loci under selection. Ecological processes, such as predator-prey interactions, could be secured by capturing intra-specific trophic variation. Additionally, different attribute spaces with different targets can be specified for different species. As long as the variation can be described using Euclidean distances, the \texttt{R} package can be used to obtain a representative sample (note that data may require transformation or dimension reduction to meet this assumption). This \texttt{R} package provides decision makers with the tools to generate prioritisations that secure both biodiversity patterns and the processes that maintain biodiversity. Additionally, the package contains functionality to accommodate uncertainty in the distribution of features, and also identify suitably connected reserves. Both the simulated and case study species suggest that conservation planning exercises need to explicitly consider biodiversity processes during the reserve selection process to ensure they are captured.

@@

The formulation requires spatially comprehensive data to map attribute spaces to planning units. For instance, in both case-studies, all planning units occupied by the species were associated with coordinates in the attribute spaces. However, most real-world data sets are patchy--some planning units will be occupied by species for which attribute space data is not available. To use such patchy data with the \texttt{raptr R} package, the gaps in the data must first be filled in. Spatially explicit models [eg. krigging or generalised dissimilarity models; @r467; @r466] could be used to estimate data for locations that are missing from the data set as a pre-processing step. This approach has been successfully applied to a range of biological data sets [eg. @r468].

To maximise the long-term persistence of biodiversity, decision makers need to identify prioritisations that preserve existing patterns of biodiversity and the processes that support them. To achieve this, conservation planners need a decision support tool that can explicitly accommodate biodiversity patterns and processes. Here, we developed the \texttt{raptr R} package to fill this void. By exploring the functionality of this package using several simulated and case study species, we found that including space-based targets can result in substantially better solutions.

@@

# Data accessibility
The \texttt{raptr R} package can be downloaded from The Comphrehensive R Archive Netowrk (\url{https://cran.r-project.org/}). All data, code, and results are stored in an online repository (\url{www.github.com/paleo13/raptr-manuscript}) to permit replication and validation of this study.

### 2016-07-25 code/rmarkdown/figures.Rmd
+ assorted manuscript fixes

---
fontsize: 11pt
documentclass: article
bibliography: references.bib
csl: reference-style.csl
output:
  rmarkdown::pdf_document:
    fig_caption: yes
    keep_tex: yes
    includes:
      in_header: [preamble-latex.tex, preamble-latex2.tex]
---

>>>>

---
fontsize: 11pt
documentclass: article
bibliography: references.bib
csl: reference-style.csl
output:
  rmarkdown::pdf_document:
    fig_caption: yes
    keep_tex: yes
    includes:
      in_header: [preamble.tex, figures-preamble.tex]
---

### 2016-07-25 code/rmarkdown/supporting-information.Rmd
+ assorted manuscript fixes

---
fontsize: 11pt
documentclass: article
bibliography: references.bib
csl: reference-style.csl
output:
  rmarkdown::pdf_document:
    fig_caption: yes
    keep_tex: yes
    includes:
      in_header: [preamble-latex.tex, preamble-latex3.tex]
---
# Supporting Information

>>>>

---
fontsize: 11pt
documentclass: article
bibliography: references.bib
csl: reference-style.csl
output:
  rmarkdown::pdf_document:
    fig_caption: yes
    keep_tex: yes
    includes:
      in_header: [preamble.tex, si-preamble.tex]
---
# Supporting Information
